### 2020-03-14 Common/chapters/medical.tex
Julien and Johan: minors

\subsection{Seriousness}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}Most cancers can be staged thanks to the TNM system. The T corresponds to the tumor size and its location; the N corresponds to whether or not the tumor has spread to draining lymph nodes; the M corresponds to the presence or absence of metastases in other parts of the body~\cite{21}. These pieces of information are used to classify cancer between four (I to IV) or sometimes five (0 to IV) different stages, reflecting the progression and the seriousness of the illness~\cite{22}. The earlier they are detected, the higher the chances of recovery are. This aspect makes cancer detection critical since every misjudgment can threaten someone's life.

@@

Other useful information such as the patient name and ID is directly stored within the DICOM files. This approach aims at linking each image to a specific person and event in order not to mix them up. Each DICOM file can be seen as part of a bigger dataset.

@@

where~$c$ is the window center,~$w$ window width, ~$P$ the pixel input value, ~$y$ the pixel output value, ~$y_{min}$ the minimal value of the output range (usually 0), ~$y_{max}$ the maximal value of the output range (usually 255). Equations \ref{eq:dicom1}, \ref{eq:dicom2} and \ref{eq:dicom3} ensure that the pixel values are correctly distributed within the output range. 
\end{enumerate}

@@

\subsection{Data format}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}Unlike the ANALYZE format that used two files to store the metadata and the actual data, the NIfTI file format stores them in one single “.nii” file, but keeps this split between the real data and the header for compatibility. This has the advantage to facilitate the use of the data and avoid storing the data without the metadata. The NIfTI format can also be compressed/decompressed on the fly using the “deflate” algorithm. A. M. Winkler affirms that this is particularly useful as "it is very common for images to have large areas of solid background, or files describing masks and regions of interest containing just a few unique values that appear repeated many time"~\cite{52}.

@@

\subsection{DICOM}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}Visualizing DICOM files is pretty straightforward since each file represents a single two-dimensional image. However, a lot of pixel transformations and normalizations have to be applied to obtain the desired result (see Section \ref{sec:dicom_data_manipulation}), which may be a source of errors. Our tool allows to display a single DICOM file as well as a sequence of files if the function is fed with a directory. Users can then scroll through the z-axis, displaying the next or previous slice.  Figure \ref{fig:visualize_lung_dcm} shows the visualization of a lung at three different z-axis coordinates. This allows the user to have a complete overview of the body part he wants to analyze.

>>>>

\subsection{Seriousness}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}Most cancers can be staged thanks to the TNM system. The T corresponds to the tumor size and its location; the N corresponds to whether or not the tumor has spread to draining lymph nodes; the M corresponds to the presence or absence of metastases in other parts of the body~\cite{21}. These pieces of information are used to classify cancer between four (I to IV) or sometimes five (0 to IV) different stages, reflecting the progression and the seriousness of the illness~\cite{22}. The earlier they are detected, the higher the chances of recovery are. This aspect makes cancer detection critical as every misjudgment can threaten someone's life.

@@

Other useful information such as the patient's name and ID is directly stored within DICOM files. This approach aims at linking each image to a specific person and event in order not to mix them up. Each DICOM file can be seen as part of a bigger dataset.

@@

where~$c$ is the window center,~$w$ window width, ~$P$ the pixel input value, \mbox{$y$ the} pixel output value, ~$y_{min}$ the minimal value of the output range (usually 0), ~$y_{max}$ the maximal value of the output range (usually 255). Equations \ref{eq:dicom1}, \ref{eq:dicom2} and \ref{eq:dicom3} ensure that the pixel values are correctly distributed within the output range. 
\end{enumerate}

@@

\subsection{Data format}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}Unlike the ANALYZE format that used two files to store the metadata and the actual data, the NIfTI file format stores them in one single “.nii” file, but keeps this split between the real data and the header for compatibility. This has the advantage to facilitate the use of the data and avoid storing the data without the metadata. The NIfTI format can also be compressed/decompressed on the fly using the “deflate” algorithm. A. M. Winkler affirms that this is particularly useful as "it is very common for images to have large areas of solid background, or files describing masks and regions of interest containing just a few unique values that appear repeated many times"~\cite{52}.

@@

\subsection{DICOM}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}Visualizing DICOM files is pretty straightforward since each file represents a single two-dimensional image. However, a lot of pixel transformations and normalizations have to be applied to obtain the desired result (see Section \ref{sec:dicom_data_manipulation}), which may be a source of errors. Our tool allows to display a single DICOM file as well as a sequence of files if the function is fed with a directory. Users can then scroll through the z-axis, displaying the next or previous slice.  Figure \ref{fig:visualize_lung_dcm} shows the visualization of a lung CT scan at three different z-axis coordinates. This allows the user to have a complete overview of the body part he wants to analyze.

### 2020-03-14 Common/chapters/paper_reproduction.tex
Julien and Johan: proofread til the end
Julien and Johan: minors

\subsection{PROSTATEx: Data processing}
\label{sec:prostatex_data_processing}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}This section describes the processing operations performed on the raw data until it passes through the neural network. All steps mentioned below are also valid for the second part of the experiment about the challenge (Section \ref{sec:classification_challenge}).

@@

\begin{algorithm}
    \caption{PROSTATEx preprocessing}
    \label{alg:PROSTATEx_preprocessing}
    \begin{algorithmic}[1] % The number tells where the line numbering should start
        \Procedure{main}{$dataset\_folder, findings\_CSV, slices\_CSV, output\_folder$}
        		\State Create output directories: $"output\_folder/True", "output\_folder/False"$\\
        		\State $findings \gets read\_CSV(findings\_CSV)$ \Comment{ProstateX-Findings-Train.csv}
        		\State $slices \gets read\_CSV(slices\_CSV)$ \Comment{ProstateX-Images-Train.csv}
			\State $meta \gets merge(findings, slices)$\Comment{Both CSV files are complementary to each other.}\\
            \For{$row$ in $meta$}
            		\State $patient\_id \gets row["ProxID"]$
            		\State $finding\_id \gets row["fid"]$
            		\State $mri\_type\_number \gets row["DCMSerNumber"]$
            		\State $clinical\_significance \gets row["ClinSig"]$
            		\State $img\_i, img\_j, img\_k \gets row["ijk"]$
            		\State $slice\_number \gets img\_k + 1$ \Comment{CSV indexing in $[0,\infty]$, DICOM in $[1,\infty]$}\\
                \For{$visit\ in\ patient\_id$'s folder}
                		\For{$mri\_type$ in $visit$}
                			\If {$mri\_type$ starts with "$mri\_type\_number$-"}
                				\For{$dicom\_file$ in $mri\_type$}
                					\If {$slice\_number == dicom\_file.InstanceNumber$}
                						\State{$slice \gets normalize\_dicom(dicom\_file)$} \Comment{see Section \ref{sec:dicom_data_manipulation}}
                						\State{Save $slice$ in $"output\_folder/clinical\_significance"$}
                					\EndIf
                				\EndFor
                			\EndIf
            			\EndFor
            		\EndFor
            \EndFor
        \EndProcedure
    \end{algorithmic}
\end{algorithm}

@@

Finally, the data was split into three subsets (training ($80$\%), validation ($10$\%) and test ($10$\%)) and was exported as NumPy arrays. The percentages represent a percentage of patients. Figure \ref{fig:paper_reproduction_split} explains the entire splitting process and gives details about the patients that are in each set. In fact, each subset contains respectively 80\%, 10\% and 10\% of all patients belonging to the "true" and "false" class. At the beginning of the process, class "false" contains $\lfloor 0.8 * 164 \rfloor = 131$ training patients, $\lfloor 0.1 * 164 \rfloor = 16$ test patients and $164 - 131 - 16 = 17$ validation patients. On the other hand, class "true" contains $\lfloor 0.8 * 69 \rfloor = 55$ training patients, $\lfloor 0.1 * 69 \rfloor = 6$ test patients and $69 - 55 - 6 = 8$ validation patients. Then, during the stacking process, some patients were not used since the three sequences were not available for them. These patients were simply discarded. At the end, the "false" class contained 131 training patients, 17 validation patients and 15 test patients, whereas the "true" class contained 55 training patients, 8 validation patients and 4 test patients. This approach does not take the number of images into account (a patient may have multiple lesions for example) but guarantees that every lesion is exclusive to a specific subset. In other words, it is impossible to find a slighty different version of the same lesion (resulting from the data augmentation) in the training set and in the validation set for example. This would be problematic since the validation set, as well as the test set, need to contain unseen data to be able to evaluate the generalization ability of the model in an unbiased way.

@@

\subsection{Data processing verification}
\label{sec:dataProcessingVerification}
\subsubsection{Cropping verification using red dots}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}In order to check the correctness of the augmentation process, images with a red dot at the lesion coordinates were generated. First of all, a red dot was placed on the full image thanks to the "i,j,k" attribute of the CSV file. This full image was exported. Then, images were augmented in the exact same manner as described in Section \ref{sec:numpyToAugmentedStacked}. The expected result satisfies the following properties:
\begin{itemize}
	\item The red dot is localized in the center of the cropped and augmented image.
	\item The three cropped and augmented images within a stacked image contain the exact same tissue. 
	\item The three cropped and augmented images within a stacked image are not rotated/shifted/flipped differently. 
	\item A lesion is visible to the naked eye. 
\end{itemize}

@@

\subsubsection{Architecture}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}The model architecture used in the paper is a modified version of the VGG network from the Oxford's Visual Geometry Group (VGG). This model was initially designed as part of the Large Scale Visual Recognition Challenge 2014 that used the ImageNet dataset composed of 14 million images belonging to 1000 classes. Figure \ref{fig:paper_model} illustrates its structure and its corresponding implementation in Python with the PyTorch framework.

It is first composed of three convolution-dropout-max-pooling blocks followed by three fully-connected-dropout blocks. Each convolutional box (in blue) in the figure represents in reality three layers: the convolutional layer, the batch normalization layer and the exponential linear unit (ELU) activation function. The same principle applies to the fully connected layer box (in orange) that is divided into a fully connected layer followed by the exponential linear unit. The last fully connected box (in purple) has the same structure except that the exponential linear unit is replaced with a softmax function for classification.

In comparison to the original VGG, this model keeps the small filter size of 3x3, also doubles the number of filters after each convolution-dropout-max-pooling block and has a stride of 1 for all convolutions. It differs from the traditional VGG in that it makes use of a smaller number of layers since the task is simpler than the original one. Moreover, it uses exponential linear units instead of rectified linear units as activation functions and adds dropout and batch normalization layers. Also, it uses 1x1 convolutions.

@@

\subsubsection{Experimental setup}

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}Keeping the exact same hyperparameters as the original paper led the experiment to disappointing results. Consequently, new hyperparameters had to be found. A grid search approach driven by the AUC obtained on the test set (the one built for the experiment, not the one of the challenge) was chosen. In practical terms, this consisted in trying multiple combinations of hyperparameters and keeping the one that produced the best AUC on this test set. At the end of this process, it appeared that the best model was obtained when the data was organized in batches of 32 samples, with an initial learning rate of $1*10^{-7}$, a learning rate scheduler on plateau using a factor of 100 with a patience of 2 (the learning rate is divided by 100 when the validation AUC decreases for two consecutive epochs) and a dropout rate of 0.2. Note that the model was initialized picking the best one among 200 models generated from a roulette maximizing the AUC. This model was implemented with PyTorch using the Adam optimizer to update its weights and was trained on an Nvidia GeForce GTX Titan X graphics card during approximately 3 hours.\\

@@

\subsection{Results}
\label{sec:paper_reproduction_results}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}Figure \ref{fig:paper_reproduction_results} shows the different metrics registered on the training and validation sets during the training phase. The best model was chosen at epoch 21 with an accuracy of $0.7525$ and an AUC of $0.765$ on the validation set (recall: the condition for a model to be chosen as best is that the current AUC and accuracy should be higher than the AUC and accuracy of the previous best model).

\noindent This best model was evaluated on the test set built for the experiment and achieved an AUC of $0.75$ using their "enhanced prediction" technique (see Figure \ref{fig:paper_reproduction_auc_on_our_test_set}).

@@

As a complement, this part of the experiment focuses on the challenge results in order to improve the AUC of $0.71$ achieved by the model trained with 80\% of all available data. In a general way, the more available data there is, the better the model performs. Consequently, the idea behind this section is to use the entire dataset as training set, discarding the validation and test sets from previous section. Apart from this new split, the data processing and the hyperparameters stay the same. The only difference in the training is about the learning rate scheduler. Since the validation set does not exist anymore, there is no possibility to use the \textit{reducelronplateau} function to prevent overfitting. However, this does not cause any problem thanks to the results of the previous section. In fact, it is clearly visible that the model would have started overfitting at epoch 17 if no learning rate correction had been applied. Hence, it is possible to conclude that, in this case, the best model should also be the one around epoch 17 (or a few epochs later). In order to ensure this, the models at epochs 15, 16, 17, 18, 19, 20, 21, 22, 23 and 30 were evaluated on the official challenge test set. For each of these models, a CSV file containing the predictions for each lesion of each patient was automatically generated and submitted to the challenge.

>>>>

\subsection{PROSTATEx: Data processing}
\label{sec:prostatex_data_processing}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}This section describes the processing operations performed on the raw data until it passes through the neural network. All the steps mentioned below are also valid for the second part of the experiment about the challenge (Section \ref{sec:classification_challenge}).

@@

\begin{algorithm}
    \caption{PROSTATEx preprocessing}
    \label{alg:PROSTATEx_preprocessing}
    \begin{algorithmic}[1] % The number tells where the line numbering should start
        \Procedure{main}{$dataset\_folder, findings\_CSV, slices\_CSV, output\_folder$}
        		\State Create output directories: $"output\_folder/True", "output\_folder/False"$\\
        		\State $findings \gets read\_CSV(findings\_CSV)$ \Comment{ProstateX-Findings-Train.csv}
        		\State $slices \gets read\_CSV(slices\_CSV)$ \Comment{ProstateX-Images-Train.csv}
			\State $meta \gets merge(findings, slices)$\Comment{Both CSV files are complementary to each other.}\\
            \For{$row$ in $meta$}
            		\State $patient\_id \gets row["ProxID"]$
            		\State $finding\_id \gets row["fid"]$
            		\State $mri\_type\_number \gets row["DCMSerNumber"]$
            		\State $clinical\_significance \gets row["ClinSig"]$
            		\State $img\_i, img\_j, img\_k \gets row["ijk"]$
            		\State $slice\_number \gets img\_k + 1$ \Comment{CSV indexing in $[0,\infty]$, DICOM in $[1,\infty]$}\\
                \For{$visit\ in\ patient\_id$'s folder}
                		\For{$mri\_type$ in $visit$}
                			\If {$mri\_type$ starts with "$mri\_type\_number$-"}
                				\For{$dicom\_file$ in $mri\_type$}
                					\If {$slice\_number == dicom\_file.InstanceNumber$}
                						\State{$slice \gets normalize\_dicom(dicom\_file)$} \Comment{see Section \ref{sec:dicom_data_manipulation}}
                						\State{Save $slice$ into $"output\_folder/clinical\_significance"$}
                					\EndIf
                				\EndFor
                			\EndIf
            			\EndFor
            		\EndFor
            \EndFor
        \EndProcedure
    \end{algorithmic}
\end{algorithm}

@@

Finally, the data was split into three subsets (training ($80$\%), validation ($10$\%) and test ($10$\%)) and was exported as NumPy arrays. The percentages represent a percentage of patients. Figure \ref{fig:paper_reproduction_split} explains the entire splitting process and gives details about the patients that are in each set. Each subset contains respectively 80\%, 10\% and 10\% of all patients belonging to the "true" and "false" classes. At the beginning of the process, class "false" contains $\lfloor 0.8 * 164 \rfloor = 131$ training patients, $\lfloor 0.1 * 164 \rfloor = 16$ test patients and $164 - 131 - 16 = 17$ validation patients. On the other hand, class "true" contains $\lfloor 0.8 * 69 \rfloor = 55$ training patients, $\lfloor 0.1 * 69 \rfloor = 6$ test patients and $69 - 55 - 6 = 8$ validation patients. Then, during the stacking process, some patients were not used since the three sequences were not available for them. These patients were simply discarded. At the end, the "false" class contained 131 training patients, 17 validation patients and 15 test patients, whereas the "true" class contained 55 training patients, 8 validation patients and 4 test patients. This approach does not take the number of images into account (a patient may have multiple lesions for example) but guarantees that every lesion is exclusive to a specific subset. In other words, it is impossible to find a slighty different version of the same lesion (resulting from the data augmentation) in the training set and in the validation set for example. This would be problematic since the validation set and the test set need to contain unseen data to be able to evaluate the generalization ability of the model in an unbiased way.

@@

\subsection{Data processing verification}
\label{sec:dataProcessingVerification}
\subsubsection{Cropping verification using red dots}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}In order to check the correctness of the augmentation process, images with a red dot at the lesion coordinates were generated. First of all, a red dot was placed on the full image thanks to the "i,j,k" attribute of the CSV file. This full image was exported. Then, images were augmented in the exact same manner as described in \mbox{Section \ref{sec:numpyToAugmentedStacked}}. The expected result satisfies the following properties:
\begin{itemize}
	\item The red dot is localized in the center of the cropped and augmented image.
	\item The three cropped and augmented images within a stacked image contain the exact same tissue. 
	\item The three cropped and augmented images within a stacked image are not rotated/shifted/flipped differently. 
	\item A lesion is visible to the naked eye. 
\end{itemize}

@@

\subsubsection{Architecture}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}The model architecture used in the paper is a modified version of the VGG network from the Oxford's Visual Geometry Group (VGG). This model was initially designed as part of the Large Scale Visual Recognition Challenge 2014 that used the ImageNet dataset composed of 14 million images belonging to 1000 classes. \mbox{Figure \ref{fig:paper_model}} illustrates its structure and its corresponding implementation in Python with the PyTorch framework.

It is first composed of three convolution-dropout-max-pooling blocks followed by three fully-connected-dropout blocks. Each convolutional box (in blue) in the figure represents in reality three layers: the convolutional layer, the batch normalization layer and the exponential linear unit (ELU) activation function. The same principle applies to the fully connected layer box (in orange) that is divided in a fully connected layer followed by the exponential linear unit. The last fully connected box (in purple) has the same structure except that the exponential linear unit is replaced with a softmax function for classification.

In comparison to the original VGG, this model keeps the small filter size of $3*3$, also doubles the number of filters after each convolution-dropout-max-pooling block and has a stride of 1 for all convolutions. It differs from the traditional VGG in that it makes use of a smaller number of layers since the task is simpler than the original one. Moreover, it uses exponential linear units instead of rectified linear units as activation functions and adds dropout and batch normalization layers. Also, it uses $1*1$ convolutions.

@@

\subsubsection{Experimental setup}
\label{sec:paper_reproduction_experimental_setup}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}Keeping the exact same hyperparameters as the original paper led the experiment to disappointing results. Consequently, new hyperparameters had to be found. A grid search approach driven by the AUC obtained on the test set (the one built for the experiment, not the one of the challenge) was chosen. In practical terms, this consisted in trying multiple combinations of hyperparameters and keeping the one that produced the best AUC on this test set. At the end of this process, it appeared that the best model was obtained when the data was organized in batches of 32 samples, with an initial learning rate of $1*10^{-7}$, a learning rate scheduler on plateau using a factor of 100 with a patience of 2 (the learning rate is divided by 100 when the validation AUC decreases for two consecutive epochs) and a dropout rate of 0.2. Note that the model was initialized picking the best one among 200 models generated from a roulette maximizing the AUC. This model was implemented with PyTorch using the Adam optimizer to update its weights and was trained on an Nvidia GeForce GTX Titan X graphics card during approximately 3 hours.\\

@@

\subsection{Results}
\label{sec:paper_reproduction_results}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}Figure \ref{fig:paper_reproduction_results} shows the different metrics registered on the training and validation sets during the training phase. The best model was chosen at epoch 21 with an accuracy of $0.7525$ and an AUC of $0.765$ on the validation set (recall: the condition for a model to be chosen as best is that the current AUC and accuracy must be higher than the AUC and accuracy of the previous best model).

\noindent This best model was evaluated on the test set built for the experiment and achieved an AUC of $0.75$ using the authors' "enhanced prediction" technique (see Figure \ref{fig:paper_reproduction_auc_on_our_test_set}).

@@

As a complement, this part of the experiment focuses on the challenge results in order to improve the AUC of $0.71$ achieved by the model trained with 80\% of all available data. In a general way, the more available data there is, the better the model performs. Consequently, the idea behind this section is to use the entire dataset as training set, discarding the validation and test sets from the previous section. Apart from this new split, the data processing and the hyperparameters stay the same. The only difference in the training is about the learning rate scheduler. Since the validation set does not exist anymore, there is no possibility to use the \textit{reducelronplateau} function to prevent overfitting. However, this does not cause any problem thanks to the results of the previous section. In fact, it is clearly visible that the model would have started overfitting at epoch 17 if no learning rate correction had been applied. Hence, it is possible to conclude that, in this case, the best model should also be the one around epoch 17 (or a few epochs later). In order to ensure this, the models at epochs 15, 16, 17, 18, 19, 20, 21, 22, 23 and 30 were evaluated on the official challenge test set. For each of these models, a CSV file containing the predictions for each lesion of each patient was automatically generated and submitted to the challenge.

### 2020-03-14 Common/chapters/transfer_learning.tex
Julien and Johan: minors

%In this case, transfer learning aims at using various small datasets in order to achieve the same or better performance than with a single huge dataset.
However, in order for the transfer learning to be useful, common low level features have to be present among the datasets on which the model is trained. Consequently, the goal of this experiment is to use transfer learning on three datasets coming from three different body parts that are prostate, brain and lung, in order to check if they share common features that could be useful to improve the performance of the model on a specific dataset. In this experiment, the model trained to classify prostate lesions is then trained to classify brain tumors and finally lung lesions. At the end of the process, the model is once again trained on the prostate dataset and the performances of the model at the beginning and at the end of the experiment are compared.

>>>>

%In this case, transfer learning aims at using various small datasets in order to achieve the same or better performance than with a single huge dataset.
However, in order for the transfer learning to be useful, common low-level features have to be present among the datasets on which the model is trained. Consequently, the goal of this experiment is to use transfer learning on three datasets coming from three different body parts that are prostate, brain and lung, in order to check if they share common features that could be useful to improve the performance of the model on a specific dataset. In this experiment, the model trained to classify prostate lesions is then trained to classify brain tumors and finally lung lesions. At the end of the process, the model is once again trained on the prostate dataset and the performances of the model at the beginning and at the end of the experiment are compared.
